---
title: "Demographic Change in Detroit, MI (Wayne County)"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

options(tigris_use_cache = TRUE)
options(tigris_progress = FALSE)

library(tidycensus)
library(tigris)
library(tmap)
library(sf)
library(dplyr)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyverse)
```

These are several analyses that I produced in the course "Spatial Demographics" taught by Billy Southern at the University of Oregon, Spring 2026. These analyses integrate information related to Census data (both ACS and 10-year) produced on either personally-chosen or other reasons.

# Introduction

I chose Detroit because...

## Population Dispersion

```{r}
options(tigris_use_cache = TRUE)
detroit_pop <- get_acs(
  geography = "tract",
  variables = "B01003_001", # population
  state = "MI",
  county = "Wayne",
  year = 2020,
  geometry = TRUE
)

ggplot(detroit_pop) +
  geom_sf(aes(fill = estimate)) +
  #scale_fill_viridis_c(labels = scales::label_dollar()) +
  scale_fill_viridis_c(option = "mako") +
  theme_void() +
  labs(fill = "Population")
```

Figure 1: ...

Figure 1 evaluates the...

## Population Pyramid

```{r}
# get data
michigan <- get_estimates(
  geography = "state",
  state = "MI",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = TRUE,
  year = 2019)

# filter out "both sexes"
michigan_filtered <- michigan %>%
  filter(str_detect(AGEGROUP, "^Age \\d+ to \\d+ years$"),
         SEX %in% c("Male", "Female")) %>%
  mutate(population = ifelse(SEX == "Male", -value, value))

# order ages
michigan_filtered <- michigan_filtered %>%
  mutate(age_start = as.numeric(str_extract(AGEGROUP, "\\d+"))) %>%
  arrange(age_start)

michigan_filtered$AGEGROUP <- factor(
  michigan_filtered$AGEGROUP,
  levels = unique(michigan_filtered$AGEGROUP))
#plot
options(scipen = 999)
michigan__pyramid <- ggplot(michigan_filtered,
                            aes(x = AGEGROUP, y = population, fill = SEX)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = abs) +
  scale_x_discrete(labels = ~ str_remove_all(.x, "Age\\s|\\syears")) +
  scale_fill_manual(values = c("#A9BC7E", "#82B5C4")) +
  labs(
    x = "Age Group",
    y = "Population",
    title = "2019 Population Structure in Michigan",
    caption = "Source: US Census Bureau"
  ) +
  theme_minimal()
michigan__pyramid
```

Figure 2: ...

Figure 2 evaluates the...

## "Young" Population Dispersion (age \< 18)

```{r}
#| echo: false
#| message: false
#| warning: false
young_detroit <- get_decennial(
  geography = "tract",
  variables = "P1_004N",
  state = "MI",
  county = "Wayne",
  geometry = TRUE,
  year = 2020
)

young_det_centroids <- st_centroid(young_detroit)

# map it
tm_shape(young_detroit) +
  tm_polygons(
    col = "value", 
    palette = "Purples", 
    style = "quantile", # Groups data evenly
    border.col = "white", 
    title = "Youth Percentage"
  )
```

Figure 3: ...

Figure 3 evaluates the...

## Dot Density: Races in Wayne County

```{r}
#| echo: false
#| message: false
#| warning: false
# dot-density maps
tmap_mode("plot")
detroit_race <- get_decennial(
  geography = "tract",
  state = "MI",
  county = "Wayne",
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N"
  ),
  summary_var = "P2_001N",
  year = 2020,
  geometry = TRUE
) %>%
  mutate(percent = 100 * (value / summary_value))

detroit_race <- detroit_race %>%
  filter(value >= 100)

detroit_dots <- detroit_race %>%
  as_dot_density(
    value = "value",
    values_per_dot = 2000,
    group = "variable"
  )

background_tracts <- filter(detroit_race, variable == "White")

tmap_mode("plot")

tm_shape(background_tracts) +
  tm_polygons() +
  tm_shape(detroit_dots) +
  tm_dots()

race_map <- tm_shape(background_tracts) +
  tm_polygons(col = "white",
              border.col = "grey") +
  tm_shape(detroit_dots) +
  tm_dots(col = "variable",
          palette = "Set1",
          size = 0.1,
          title = "1 dot = 100 people") +
  tm_layout(title = "",
            legend.position = c("right","bottom"))
race_map
```

Figure 5: Race ethnicity (2020 Census)

In Figure 5, we see
