---
title: "Demographic Change in Detroit, MI (Wayne County)"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

options(tigris_use_cache = TRUE)
options(tigris_progress = FALSE)

library(tidycensus)
library(tigris)
library(tmap)
library(sf)
library(dplyr)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyverse)
library(sf)
library(rmapshaper)
library(flextable)
library(segregation)

tmap_mode("plot")
```

These are several analyses that I produced in the course "Spatial Demographics" taught by Billy Southern at the University of Oregon, Spring 2026. These analyses integrate information related to Census data (both ACS and 10-year) produced on either personally-chosen or other reasons.

# Introduction

I chose Detroit because...

## Population Dispersion

```{r}
options(tigris_use_cache = TRUE)
detroit_pop <- get_acs(
  geography = "tract",
  variables = "B01003_001", # population
  state = "MI",
  county = "Wayne",
  year = 2020,
  geometry = TRUE
)

ggplot(detroit_pop) +
  geom_sf(aes(fill = estimate)) +
  #scale_fill_viridis_c(labels = scales::label_dollar()) +
  scale_fill_viridis_c(option = "mako") +
  theme_void() +
  labs(fill = "Population")
```

Figure 1: ...

Figure 1 evaluates the...

## Population Pyramid (State of Michigan)

```{r}
# get data
michigan <- get_estimates(
  geography = "state",
  state = "MI",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = TRUE,
  year = 2019)

# filter out "both sexes"
michigan_filtered <- michigan %>%
  filter(str_detect(AGEGROUP, "^Age \\d+ to \\d+ years$"),
         SEX %in% c("Male", "Female")) %>%
  mutate(population = ifelse(SEX == "Male", -value, value))

# order ages
michigan_filtered <- michigan_filtered %>%
  mutate(age_start = as.numeric(str_extract(AGEGROUP, "\\d+"))) %>%
  arrange(age_start)

michigan_filtered$AGEGROUP <- factor(
  michigan_filtered$AGEGROUP,
  levels = unique(michigan_filtered$AGEGROUP))
#plot
options(scipen = 999)
michigan__pyramid <- ggplot(michigan_filtered,
                            aes(x = AGEGROUP, y = population, fill = SEX)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = abs) +
  scale_x_discrete(labels = ~ str_remove_all(.x, "Age\\s|\\syears")) +
  scale_fill_manual(values = c("#A9BC7E", "#82B5C4")) +
  labs(
    x = "Age Group",
    y = "Population",
    title = "2019 Population Structure in Michigan",
    caption = "Source: US Census Bureau"
  ) +
  theme_minimal()
michigan__pyramid
```

Figure 2: ...

Figure 2 evaluates the...

## "Young" Population Dispersion (age \< 18)

```{r}
#| echo: false
#| message: false
#| warning: false
young_detroit <- get_decennial(
  geography = "tract",
  variables = "P1_004N",
  state = "MI",
  county = "Wayne",
  geometry = TRUE,
  year = 2020
)

young_det_centroids <- st_centroid(young_detroit)

# map it
tm_shape(young_detroit) +
  tm_polygons(
    col = "value", 
    palette = "Purples", 
    style = "quantile", # Groups data evenly
    border.col = "white", 
    title = "Youth Percentage"
  )
```

Figure 3: ...

Figure 3 evaluates the...

## Race Distribution

### Dot Density

```{r}
#| echo: false
#| message: false
#| warning: false
# dot-density maps
tmap_mode("plot")
detroit_race <- get_decennial(
  geography = "tract",
  state = "MI",
  county = "Wayne",
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N"
  ),
  summary_var = "P2_001N",
  year = 2020,
  geometry = TRUE
) %>%
  mutate(percent = 100 * (value / summary_value))

detroit_race <- detroit_race %>%
  filter(value >= 100)

detroit_dots <- detroit_race %>%
  as_dot_density(
    value = "value",
    values_per_dot = 2000,
    group = "variable"
  )

background_tracts <- filter(detroit_race, variable == "White")

tmap_mode("plot")

tm_shape(background_tracts) +
  tm_polygons() +
  tm_shape(detroit_dots) +
  tm_dots()

race_map <- tm_shape(background_tracts) +
  tm_polygons(col = "white",
              border.col = "grey") +
  tm_shape(detroit_dots) +
  tm_dots(col = "variable",
          palette = "Set1",
          size = 0.1,
          title = "1 dot = 100 people") +
  tm_layout(title = "",
            legend.position = c("right","bottom"))
race_map
```

Figure 4: Race ethnicity (2020 Census)

In Figure 4, we see

### Faceted

```{r}
tmap_mode("plot")

detroit_race <- get_decennial(
  geography = "tract",
  state = "MI",
  county = "Wayne",
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N"
  ),
  summary_var = "P2_001N",
  year = 2020,
  geometry = TRUE
) %>%
  mutate(percent = 100 * (value / summary_value))

tm_shape(detroit_race) +
  tm_facets(by = "variable", ncol = 2, nrow = 3, scale.factor = 4) +
  tm_fill(
    col = "percent",
    style = "quantile",
    n = 6,
    palette = "Blues",
    title = "Percent (2020 US Census)",
    showNA = FALSE
  ) +
  tm_layout(
    bg.color = "grey",
    panel.label.bg.color = "white",
    legend.outside = TRUE,
    legend.outside.position = "right"
  )
```

Figure 5: Race ethnicity (2020 Census)

In Figure 5, we see

## Visualizing Segregation and Gentrification

### Multigroup Segregation Index

```{r}
#| echo: false
#| message: false
#| warning: false
# Bring in 2019-2023 census tract data using the Census API
dtown.tracts <- get_acs(geography = "tract",
                     year = 2023,
                     variables = c(tpop = "B03002_001",
                                   white = "B03002_003", black = "B03002_004",
                                   asian = "B03002_006", hisp = "B03002_012"),
                     state = "MI",
                     survey = "acs5",
                     output = "wide",
                     geometry = TRUE)
# Calculate, rename and keep essential vars.
dtown.tracts <- dtown.tracts %>%
  mutate(pwhite = 100*(whiteE/tpopE), pasian = 100*(asianE/tpopE),
         pblack = 100*(blackE/tpopE), phisp = 100*(hispE/tpopE)) %>%
  rename(white = whiteE, asian = asianE, black = blackE,
         hisp = hispE, tpop = tpopE) %>%
  select(GEOID,tpop, pwhite, pasian, pblack, phisp,
         white, asian, black, hisp)
# Bring in city boundaries
mi_pl <- places(state = "MI", year = 2023, cb = TRUE)
# Keep four large cities in CA
large.cities <- mi_pl %>%
  filter(NAME == "Detroit" |
           NAME == "Grand Rapids" | NAME == "Ann Arbor" |
           NAME == "Lansing")
#Clip tracts in large cities
large.tracts <- ms_clip(target = dtown.tracts,
                        clip = large.cities,
                        remove_slivers = TRUE)

# glimpse(large.tracts)
# names(large.tracts)
# names(large.cities)

sf::sf_use_s2(FALSE)
large.tracts <- large.tracts %>%
  st_join(large.cities)
sf::sf_use_s2(TRUE)
#names(large.tracts) # lots of new variables that we don't need

# remove duplicates (ending in .x and .y)
large.tracts <- large.tracts %>%
  select(-(STATEFP:GEOIDFQ), -(NAMELSAD:AWATER))
#names(large.tracts) # check

hisp_map <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  tm_shape() +
  tm_polygons(fill = "phisp",
              fill.scale = tm_scale(style = "quantile", values = "reds"),
              frame = FALSE) +
  tm_title("Hispanic")

black_map <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  tm_shape() +
  tm_polygons(fill = "pblack",
              fill.scale = tm_scale(style = "quantile", values = "reds"),
              frame = FALSE) +
  tm_title("Black")

asian_map <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  tm_shape() +
  tm_polygons(fill = "pasian",
              fill.scale = tm_scale(style = "quantile", values = "reds"),
              frame = FALSE) +
  tm_title("Asian")

white_map <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  tm_shape() +
  tm_polygons(fill = "pwhite",
              fill.scale = tm_scale(style = "quantile", values = "reds"),
              frame = FALSE) +
  tm_title("White")

tmap_arrange(
  hisp_map, black_map,
  asian_map, white_map,
  ncol = 2
)
```

Figure 6:

### Dissimilarity Index:

```{r}
#| echo: false
#| message: false
#| warning: false

library(sf)

# DROP geometry BEFORE ANY GROUPING
tract_data <- large.tracts %>%
  st_drop_geometry()

# totals
tract_data <- tract_data %>%
  group_by(NAME) %>%
  mutate(
    whitec = sum(white),
    asianc = sum(asian),
    blackc = sum(black),
    hispc  = sum(hisp)
  ) %>%
  ungroup()

# contributions
tract_data <- tract_data %>%
  mutate(
    d.wb = abs(black/blackc - white/whitec),
    d.wa = abs(asian/asianc - white/whitec),
    d.wh = abs(hisp/hispc - white/whitec)
  )

# final index (NO sf involved anymore)
dissimilarity <- tract_data %>%
  group_by(NAME) %>%
  summarize(
    BWD = 0.5 * sum(d.wb, na.rm = TRUE),
    AWD = 0.5 * sum(d.wa, na.rm = TRUE),
    HWD = 0.5 * sum(d.wh, na.rm = TRUE)
  )

dis.table <- flextable(dissimilarity) %>%
  colformat_double(j = c("BWD", "AWD", "HWD"), digits = 3)

dis.table
```

Table 1: Dissimilarity Index betwe....

Table 1 shows us that...

### Interaction Index (P\*)

```{r}
#| echo: false
#| message: false
#| warning: false

large.tracts <- large.tracts %>%
  group_by(NAME) %>%
  mutate(
    whitec = sum(white),
    asianc = sum(asian),
    blackc = sum(black),
    hispc  = sum(hisp),
    tpopc  = sum(tpop)
  ) %>%
  ungroup()

large.tracts <- large.tracts %>%
  mutate(
    i.wb = (black / blackc) * (white / tpop),
    i.wa = (asian / asianc) * (white / tpop),
    i.wh = (hisp / hispc) * (white / tpop)
  )

interaction_summary <- large.tracts %>%
  st_drop_geometry() %>%
  summarize(
    BWI = sum(i.wb, na.rm = TRUE),
    AWI = sum(i.wa, na.rm = TRUE),
    HWI = sum(i.wh, na.rm = TRUE)
  )

int.table <- flextable(interaction_summary) %>%
  colformat_double(j = c("BWI", "AWI", "HWI"), digits = 3)

int.table
```

Table 2:

Table 2 shows us...

### Location Quotient

```{r}
# 1. Create LQ values (sf)
large.tracts <- large.tracts %>%
  group_by(NAME) %>%
  mutate(
    whitec = sum(white),
    asianc = sum(asian),
    blackc = sum(black),
    hispc  = sum(hisp),
    tpopc  = sum(tpop)
  ) %>%
  ungroup()

large.tracts <- large.tracts %>%
  mutate(
    i.wb = (black / blackc) * (white / tpop),
    i.wa = (asian / asianc) * (white / tpop),
    i.wh = (hisp / hispc) * (white / tpop)
  )

dtown.tracts <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  mutate(
    blklq   = (black / tpop) / (blackc / tpopc),
    asnlq   = (asian / tpop) / (asianc / tpopc),
    hisplq  = (hisp / tpop) / (hispc / tpopc),
    whitelq = (white / tpop) / (whitec / tpopc)
  )

# 2. Histogram (no geometry)
ggplot(st_drop_geometry(dtown.tracts), aes(blklq)) +
  geom_histogram(bins = 30)
```

Figure 7:

Figure 7 tells us that...

```{r}
#| echo: false
#| message: false
#| warning: false

library(tmap)
library(dplyr)
library(tidyr)
library(sf)

# ----------------------------
# 1. Ensure correct dataset
# ----------------------------
lq_data <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  st_as_sf()

# ----------------------------
# 2. Compute city totals ONCE
# ----------------------------
lq_data <- lq_data %>%
  mutate(
    whitec = sum(white, na.rm = TRUE),
    blackc = sum(black, na.rm = TRUE),
    asianc = sum(asian, na.rm = TRUE),
    hispc  = sum(hisp,  na.rm = TRUE),
    tpopc  = sum(tpop,  na.rm = TRUE)
  )

# ----------------------------
# 3. Compute Location Quotients
# ----------------------------
lq_data <- lq_data %>%
  mutate(
    blklq   = (black / tpop) / (blackc / tpopc),
    asnlq   = (asian / tpop) / (asianc / tpopc),
    hisplq  = (hisp  / tpop) / (hispc  / tpopc),
    whitelq = (white / tpop) / (whitec / tpopc)
  )

# ----------------------------
# 4. Pivot to long format for faceting
# ----------------------------
lq_long <- lq_data %>%
  select(geometry, blklq, asnlq, hisplq, whitelq) %>%
  pivot_longer(
    cols = c(blklq, asnlq, hisplq, whitelq),
    names_to = "race",
    values_to = "lq"
  )

# ----------------------------
# 5. Plot (2x2 layout)
# ----------------------------
tmap_mode("plot")

tm_shape(lq_long) +
  tm_polygons(
    fill = "lq",
    fill.scale = tm_scale(style = "quantile", n = 6),
    col_alpha = 0
  ) +
  tm_facets(by = "race", ncol = 2) +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE,
    title = "Location Quotients (Detroit)"
  )
```

Figure 8:

Figure 8 shows...

## Final Segregation Index

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: hide
library(tidycensus)
library(tidyverse)
library(segregation)
library(tigris)
library(sf)

# Get Michigan tract data by race/ethnicity
mi_acs_data <- get_acs(
  geography = "tract",
  variables = c(
    white = "B03002_003",
    black = "B03002_004",
    asian = "B03002_006",
    hispanic = "B03002_012"
  ),
  state = "MI",
  geometry = TRUE,
  year = 2019)

# Use tidycensus to get urbanized areas by population with geometry,
# then filter for those that have populations of 750,000 or more
us_urban_areas <- get_acs(
  geography = "urban area",
  variables = "B01001_001",
  geometry = TRUE,
  year = 2019,
  survey = "acs1"
) %>%
  filter(estimate >= 750000) %>%
  transmute(urban_name = str_remove(NAME,
                                    fixed(", MI Urbanized Area (2010)")))
# Compute an inner spatial join between the California tracts and the
# urbanized areas, returning tracts in the largest California urban
# areas with the urban_name column appended
mi_urban_data <- mi_acs_data %>%
  st_join(us_urban_areas, left = FALSE) %>%
  select(-NAME) %>%
  st_drop_geometry()


# Dissimilarity index: D
mi_urban_data %>%
  filter(variable %in% c("white", "hispanic"),
         urban_name == "Detroit") %>% dissimilarity(
           group = "variable",
           unit = "GEOID",
           weight = "estimate")
mi_urban_data %>%
  filter(variable %in% c("white", "hispanic")) %>%
  group_by(urban_name) %>%
  group_modify(~
                 dissimilarity(.x,
                               group = "variable",
                               unit = "GEOID",
                               weight = "estimate"
                 )) %>%
  arrange(desc(est))

# multi-group segregation indices
mutual_within(
  data = mi_urban_data,
  group = "variable",
  unit = "GEOID",
  weight = "estimate",
  within = "urban_name",
  wide = TRUE)
detroit_local_seg <- mi_urban_data %>%
  filter(urban_name == "Detroit") %>%
  mutual_local(
    group = "variable",
    unit = "GEOID",
    weight = "estimate",
    wide = TRUE
  )
detroit_tracts_seg <- tracts("MI", cb = TRUE, year = 2019) %>%
  inner_join(detroit_local_seg, by = "GEOID")
detroit_tracts_seg %>%
  ggplot(aes(fill = ls)) + #
  geom_sf(color = NA) +
  coord_sf(crs = 26946) +
  scale_fill_viridis_c(option = "mako", direction = -1) +
  theme_void() +
  labs(fill = "Local\nsegregation index")
```

Figure 9:
