---
title: "Demographic Change in Detroit, MI (Wayne County)"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

options(tigris_use_cache = TRUE)
options(tigris_progress = FALSE)

library(tidycensus)
library(tigris)
library(tmap)
library(sf)
library(dplyr)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyverse)
library(sf)
library(rmapshaper)
library(flextable)
library(segregation)

tmap_mode("plot")
```

These are several analyses that I produced in the course "Spatial Demographics" taught by Billy Southern at the University of Oregon, Spring 2026. These analyses integrate information related to Census data (both ACS and 10-year).

# Introduction

I chose Detroit because it's a city I've been familiar with my entire life: D-Town! More importantly, though, Detroit has an impressive history of segregation, enforced through the system: discriminatory housing covenants, redlining, and physical barriers. Federal politics sustained this structural inequality, and although many practices have passed, the memories of a segregated community thrive.

-   **Restrictive Covenants:** By 1940, roughly 80% of Detroit's residential properties had legally binding contracts that allowed African Americans from owning or occupying homes.

-   **Redlining:** The Home Owners' Loan Corporation (HOLC) and Federal Housing Administration (FHA) marked neighborhoods with large Black populations as "hazardous," denying them mortgages and home improvement loans.

-   **The Birwood Wall:** In 1941, to secure federal funding for an all-white housing development, a developer built a six-foot concrete wall to separate new white homes from the neighboring Black community.

-   **Highway Construction:** in the 1950s and 1960s, city planners routed the Chrysler and I-75 highways directly through the heart of thriving Black districts (Paradise Valley, Black Bottom). This displaced thousands, and destroyed their wealth.

-   **The Sojourner Truth Riots (1942):** When the federal government attempted to open a housing project for Black war workers in a white neighborhood, the KKK and violent mobs protested, resulting in police arresting hundreds of *Black* residents.

-   **1943, 1967 Riots:** Decades of severe overcrowding, police brutality, and housing segregation resulted in boiling social tensions and major rebellions in these years.

-   Following the 1967 civil unrest, there was a massive out-migration of white residents to the suburbs, leaving the city center with a heavily concentrated Black population (and depleting much of its tax base).

-   **School Segregation (1974):** Civil rights advocates pushed for a metro-wide school bus system to desegregate the city and suburbs. However, the U.S. Supreme Court ruled this down in *Milliken v. Bradley*, ensuring that Detroit's heavily Black school system remained completely separated from predominantly white suburban districts.

## Population Dispersion

```{r}
options(tigris_use_cache = TRUE)
detroit_pop <- get_acs(
  geography = "tract",
  variables = "B01003_001", # population
  state = "MI",
  county = "Wayne",
  year = 2020,
  geometry = TRUE
)

ggplot(detroit_pop) +
  geom_sf(aes(fill = estimate)) +
  #scale_fill_viridis_c(labels = scales::label_dollar()) +
  scale_fill_viridis_c(option = "mako") +
  theme_void() +
  labs(fill = "Population")
```

This data is from the 1-year ACS survey (2019). Population is organized by census tract—small, relatively permanent subdivisions of counties defined by the U.S. Census Bureau to contain between 1,200 and 8,000 residents. There are 297 tracts in Detroit, Michigan, and several hundred more in the entirety of Wayne County, MI.

## Population Pyramid (State of Michigan)

```{r}
# get data
michigan <- get_estimates(
  geography = "state",
  state = "MI",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = TRUE,
  year = 2019)

# filter out "both sexes"
michigan_filtered <- michigan %>%
  filter(str_detect(AGEGROUP, "^Age \\d+ to \\d+ years$"),
         SEX %in% c("Male", "Female")) %>%
  mutate(population = ifelse(SEX == "Male", -value, value))

# order ages
michigan_filtered <- michigan_filtered %>%
  mutate(age_start = as.numeric(str_extract(AGEGROUP, "\\d+"))) %>%
  arrange(age_start)

michigan_filtered$AGEGROUP <- factor(
  michigan_filtered$AGEGROUP,
  levels = unique(michigan_filtered$AGEGROUP))
#plot
options(scipen = 999)
michigan__pyramid <- ggplot(michigan_filtered,
                            aes(x = AGEGROUP, y = population, fill = SEX)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = abs) +
  scale_x_discrete(labels = ~ str_remove_all(.x, "Age\\s|\\syears")) +
  scale_fill_manual(values = c("#A9BC7E", "#82B5C4")) +
  labs(
    x = "Age Group",
    y = "Population",
    title = "2019 Population Structure in Michigan",
    caption = "Source: US Census Bureau"
  ) +
  theme_minimal()
michigan__pyramid
```

This data is also from the 2019 ACS survey (2019). Population is organized by four-year age groups—big enough to avoid 84 rows on this pyramid, but small enough to still visualize the dispersion of ages.

## "Young" Population Dispersion (age \< 18)

```{r}
#| echo: false
#| message: false
#| warning: false
young_detroit <- get_decennial(
  geography = "tract",
  variables = "P1_004N",
  state = "MI",
  county = "Wayne",
  geometry = TRUE,
  year = 2020
)

young_det_centroids <- st_centroid(young_detroit)

# map it
tm_shape(young_detroit) +
  tm_polygons(
    col = "value", 
    palette = "Purples", 
    style = "quantile", # Groups data evenly
    border.col = "white", 
    title = "Youth Percentage"
  )
```

## Race Distribution

Data taken from the 10-year United States Census (2020). Each dot represents 2000 people.

### Dot Density

```{r}
#| echo: false
#| message: false
#| warning: false
# dot-density maps
tmap_mode("plot")
detroit_race <- get_decennial(
  geography = "tract",
  state = "MI",
  county = "Wayne",
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N"
  ),
  summary_var = "P2_001N",
  year = 2020,
  geometry = TRUE
) %>%
  mutate(percent = 100 * (value / summary_value))

detroit_race <- detroit_race %>%
  filter(value >= 100)

detroit_dots <- detroit_race %>%
  as_dot_density(
    value = "value",
    values_per_dot = 2000,
    group = "variable"
  )

background_tracts <- filter(detroit_race, variable == "White")

tmap_mode("plot")

tm_shape(background_tracts) +
  tm_polygons() +
  tm_shape(detroit_dots) +
  tm_dots()

race_map <- tm_shape(background_tracts) +
  tm_polygons(col = "white",
              border.col = "grey") +
  tm_shape(detroit_dots) +
  tm_dots(col = "variable",
          palette = "Set1",
          size = 0.1,
          title = "1 dot = 100 people") +
  tm_layout(title = "",
            legend.position = c("right","bottom"))
race_map
```

### Faceted

Density of five main races in Detroit: Hispanic, Black, Native American, White, and Asian.

```{r}
tmap_mode("plot")

detroit_race <- get_decennial(
  geography = "tract",
  state = "MI",
  county = "Wayne",
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N"
  ),
  summary_var = "P2_001N",
  year = 2020,
  geometry = TRUE
) %>%
  mutate(percent = 100 * (value / summary_value))

tm_shape(detroit_race) +
  tm_facets(by = "variable", ncol = 2, nrow = 3, scale.factor = 4) +
  tm_fill(
    col = "percent",
    style = "quantile",
    n = 6,
    palette = "Blues",
    title = "Percent (2020 US Census)",
    showNA = FALSE
  ) +
  tm_layout(
    bg.color = "grey",
    panel.label.bg.color = "white",
    legend.outside = TRUE,
    legend.outside.position = "right"
  )
```

## Visualizing Segregation and Gentrification

### Multi-group Segregation Index

```{r}
#| echo: false
#| message: false
#| warning: false
# Bring in 2019-2023 census tract data using the Census API
dtown.tracts <- get_acs(geography = "tract",
                     year = 2023,
                     variables = c(tpop = "B03002_001",
                                   white = "B03002_003", black = "B03002_004",
                                   asian = "B03002_006", hisp = "B03002_012"),
                     state = "MI",
                     survey = "acs5",
                     output = "wide",
                     geometry = TRUE)
# Calculate, rename and keep essential vars.
dtown.tracts <- dtown.tracts %>%
  mutate(pwhite = 100*(whiteE/tpopE), pasian = 100*(asianE/tpopE),
         pblack = 100*(blackE/tpopE), phisp = 100*(hispE/tpopE)) %>%
  rename(white = whiteE, asian = asianE, black = blackE,
         hisp = hispE, tpop = tpopE) %>%
  select(GEOID,tpop, pwhite, pasian, pblack, phisp,
         white, asian, black, hisp)
# Bring in city boundaries
mi_pl <- places(state = "MI", year = 2023, cb = TRUE)
# Keep four large cities in CA
large.cities <- mi_pl %>%
  filter(NAME == "Detroit" |
           NAME == "Grand Rapids" | NAME == "Ann Arbor" |
           NAME == "Lansing")
#Clip tracts in large cities
large.tracts <- ms_clip(target = dtown.tracts,
                        clip = large.cities,
                        remove_slivers = TRUE)

# glimpse(large.tracts)
# names(large.tracts)
# names(large.cities)

sf::sf_use_s2(FALSE)
large.tracts <- large.tracts %>%
  st_join(large.cities)
sf::sf_use_s2(TRUE)
#names(large.tracts) # lots of new variables that we don't need

# remove duplicates (ending in .x and .y)
large.tracts <- large.tracts %>%
  select(-(STATEFP:GEOIDFQ), -(NAMELSAD:AWATER))
#names(large.tracts) # check

hisp_map <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  tm_shape() +
  tm_polygons(fill = "phisp",
              fill.scale = tm_scale(style = "quantile", values = "reds"),
              frame = FALSE) +
  tm_title("Hispanic")

black_map <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  tm_shape() +
  tm_polygons(fill = "pblack",
              fill.scale = tm_scale(style = "quantile", values = "reds"),
              frame = FALSE) +
  tm_title("Black")

asian_map <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  tm_shape() +
  tm_polygons(fill = "pasian",
              fill.scale = tm_scale(style = "quantile", values = "reds"),
              frame = FALSE) +
  tm_title("Asian")

white_map <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  tm_shape() +
  tm_polygons(fill = "pwhite",
              fill.scale = tm_scale(style = "quantile", values = "reds"),
              frame = FALSE) +
  tm_title("White")

tmap_arrange(
  hisp_map, black_map,
  asian_map, white_map,
  ncol = 2
)
```

### Dissimilarity Index:

```{r}
#| echo: false
#| message: false
#| warning: false

library(sf)

# DROP geometry BEFORE ANY GROUPING
tract_data <- large.tracts %>%
  st_drop_geometry()

# totals
tract_data <- tract_data %>%
  group_by(NAME) %>%
  mutate(
    whitec = sum(white),
    asianc = sum(asian),
    blackc = sum(black),
    hispc  = sum(hisp)
  ) %>%
  ungroup()

# contributions
tract_data <- tract_data %>%
  mutate(
    d.wb = abs(black/blackc - white/whitec),
    d.wa = abs(asian/asianc - white/whitec),
    d.wh = abs(hisp/hispc - white/whitec)
  )

# final index (NO sf involved anymore)
dissimilarity <- tract_data %>%
  group_by(NAME) %>%
  summarize(
    BWD = 0.5 * sum(d.wb, na.rm = TRUE),
    AWD = 0.5 * sum(d.wa, na.rm = TRUE),
    HWD = 0.5 * sum(d.wh, na.rm = TRUE)
  )

dis.table <- flextable(dissimilarity) %>%
  colformat_double(j = c("BWD", "AWD", "HWD"), digits = 3)

dis.table
```

Dissimilarity Indices are widely used demographic metrics that measure the evennes with which two distinct groups are distributed across a geography. It ranges from 0 to 1, where a 0 = complete integration and 1 = complete segregation.

### Interaction Index (P\*)

```{r}
#| echo: false
#| message: false
#| warning: false

large.tracts <- large.tracts %>%
  group_by(NAME) %>%
  mutate(
    whitec = sum(white),
    asianc = sum(asian),
    blackc = sum(black),
    hispc  = sum(hisp),
    tpopc  = sum(tpop)
  ) %>%
  ungroup()

large.tracts <- large.tracts %>%
  mutate(
    i.wb = (black / blackc) * (white / tpop),
    i.wa = (asian / asianc) * (white / tpop),
    i.wh = (hisp / hispc) * (white / tpop)
  )

interaction_summary <- large.tracts %>%
  st_drop_geometry() %>%
  summarize(
    BWI = sum(i.wb, na.rm = TRUE),
    AWI = sum(i.wa, na.rm = TRUE),
    HWI = sum(i.wh, na.rm = TRUE)
  )

int.table <- flextable(interaction_summary) %>%
  colformat_double(j = c("BWI", "AWI", "HWI"), digits = 3)

int.table
```

The interaction index is a metric used to measure the probability of contact or exposure between two distinct groups (such as demographic populations) across residential or social zones.

## Location Quotient

```{r}
# 1. Create LQ values (sf)
large.tracts <- large.tracts %>%
  group_by(NAME) %>%
  mutate(
    whitec = sum(white),
    asianc = sum(asian),
    blackc = sum(black),
    hispc  = sum(hisp),
    tpopc  = sum(tpop)
  ) %>%
  ungroup()

large.tracts <- large.tracts %>%
  mutate(
    i.wb = (black / blackc) * (white / tpop),
    i.wa = (asian / asianc) * (white / tpop),
    i.wh = (hisp / hispc) * (white / tpop)
  )

dtown.tracts <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  mutate(
    blklq   = (black / tpop) / (blackc / tpopc),
    asnlq   = (asian / tpop) / (asianc / tpopc),
    hisplq  = (hisp / tpop) / (hispc / tpopc),
    whitelq = (white / tpop) / (whitec / tpopc)
  )

# 2. Histogram (no geometry)
dtown_long <- dtown.tracts %>%
  st_drop_geometry() %>%
  select(blklq, asnlq, hisplq, whitelq) %>%
  pivot_longer(
    cols = everything(),
    names_to = "race",
    values_to = "lq"
  ) %>%
  # Clean up the labels and set the order for the 2x2 grid
  mutate(race = factor(race, 
                       levels = c("whitelq", "blklq", "hisplq", "asnlq"),
                       labels = c("White Population LQ", "Black Population LQ", 
                                  "Hispanic Population LQ", "Asian Population LQ")))

# 2. Plot using facet_wrap
ggplot(dtown_long, aes(x = lq)) +
  geom_histogram(bins = 30, fill = "#6596C9", color = "black", alpha = 0.7) +
  # nrow = 2 and ncol = 2 forces the 2x2 grid layout
  facet_wrap(~race, nrow = 2, ncol = 2, scales = "free_y") + 
  labs(
    x = "Location Quotient (LQ)", 
    y = "Count",
    title = ""
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 11), # Makes facet titles bold
    panel.spacing = unit(1, "lines")                   # Adds a bit of breathing room between plots
  )
```

The location quotient is a statistical ratio used in economic geography to measure how concentrated an industry, occupation, or demographic is in a specific region compared to a broader benchmark (i.e., the national average). If LQ = 1.0, the region is equally concentrated in that industry compared to the nation; it is economically diverse. If LQ \> 1.0, the industry is more concentrated in the region than in the nation. If LQ \< 1.0, the industry is less concentrated locally than at the national level

```{r}
#| echo: false
#| message: false
#| warning: false

library(tmap)
library(dplyr)
library(tidyr)
library(sf)

# ----------------------------
# 1. Ensure correct dataset
# ----------------------------
lq_data <- large.tracts %>%
  filter(NAME == "Detroit") %>%
  st_as_sf()

# ----------------------------
# 2. Compute city totals ONCE
# ----------------------------
lq_data <- lq_data %>%
  mutate(
    whitec = sum(white, na.rm = TRUE),
    blackc = sum(black, na.rm = TRUE),
    asianc = sum(asian, na.rm = TRUE),
    hispc  = sum(hisp,  na.rm = TRUE),
    tpopc  = sum(tpop,  na.rm = TRUE)
  )

# ----------------------------
# 3. Compute Location Quotients
# ----------------------------
lq_data <- lq_data %>%
  mutate(
    blklq   = (black / tpop) / (blackc / tpopc),
    asnlq   = (asian / tpop) / (asianc / tpopc),
    hisplq  = (hisp  / tpop) / (hispc  / tpopc),
    whitelq = (white / tpop) / (whitec / tpopc)
  )

# ----------------------------
# 4. Pivot to long format for faceting
# ----------------------------
lq_long <- lq_data %>%
  select(geometry, blklq, asnlq, hisplq, whitelq) %>%
  pivot_longer(
    cols = c(blklq, asnlq, hisplq, whitelq),
    names_to = "race",
    values_to = "lq"
  ) %>%
  # NEW: Convert the "race" column to a factor with human-readable labels
  mutate(race = factor(race, 
                       levels = c("whitelq", "blklq", "hisplq", "asnlq"),
                       labels = c("White LQ", "Black LQ", "Hispanic LQ", "Asian LQ")))

# ----------------------------
# 5. Plot (2x2 layout)
# ----------------------------
tmap_mode("plot")

tm_shape(lq_long) +
  tm_polygons(
    fill = "lq",
    fill.scale = tm_scale(style = "quantile", n = 6, label = "Location Quotient"), # name legend
    col_alpha = 0
  ) +
  tm_facets(by = "race", ncol = 2) +
  tm_layout(
    frame = FALSE,
    legend.outside = FALSE,
    title = ""
  )
```

## Final Segregation Index

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: hide
library(tidycensus)
library(tidyverse)
library(segregation)
library(tigris)
library(sf)

# Get Michigan tract data by race/ethnicity
mi_acs_data <- get_acs(
  geography = "tract",
  variables = c(
    white = "B03002_003",
    black = "B03002_004",
    asian = "B03002_006",
    hispanic = "B03002_012"
  ),
  state = "MI",
  geometry = TRUE,
  year = 2019)

# Use tidycensus to get urbanized areas by population with geometry,
# then filter for those that have populations of 750,000 or more
us_urban_areas <- get_acs(
  geography = "urban area",
  variables = "B01001_001",
  geometry = TRUE,
  year = 2019,
  survey = "acs1"
) %>%
  filter(estimate >= 750000) %>%
  transmute(urban_name = str_remove(NAME,
                                    fixed(", MI Urbanized Area (2010)")))
# Compute an inner spatial join between the California tracts and the
# urbanized areas, returning tracts in the largest California urban
# areas with the urban_name column appended
mi_urban_data <- mi_acs_data %>%
  st_join(us_urban_areas, left = FALSE) %>%
  select(-NAME) %>%
  st_drop_geometry()


# Dissimilarity index: D
mi_urban_data %>%
  filter(variable %in% c("white", "hispanic"),
         urban_name == "Detroit") %>% dissimilarity(
           group = "variable",
           unit = "GEOID",
           weight = "estimate")
mi_urban_data %>%
  filter(variable %in% c("white", "hispanic")) %>%
  group_by(urban_name) %>%
  group_modify(~
                 dissimilarity(.x,
                               group = "variable",
                               unit = "GEOID",
                               weight = "estimate"
                 )) %>%
  arrange(desc(est))

# multi-group segregation indices
mutual_within(
  data = mi_urban_data,
  group = "variable",
  unit = "GEOID",
  weight = "estimate",
  within = "urban_name",
  wide = TRUE)
detroit_local_seg <- mi_urban_data %>%
  filter(urban_name == "Detroit") %>%
  mutual_local(
    group = "variable",
    unit = "GEOID",
    weight = "estimate",
    wide = TRUE
  )
detroit_tracts_seg <- tracts("MI", cb = TRUE, year = 2019) %>%
  inner_join(detroit_local_seg, by = "GEOID")
detroit_tracts_seg %>%
  ggplot(aes(fill = ls)) + #
  geom_sf(color = NA) +
  coord_sf(crs = 26946) +
  scale_fill_viridis_c(option = "mako", direction = -1) +
  theme_void() +
  labs(fill = "Local\nsegregation index")
```
